Alternative Methods for Retaining Explicit and Finding Implicit Sharing in Embedded DSLs

### 2023-02-16 paper/paper.tex
last minute updates to paper

% This is samplepaper.tex, a sample chapter demonstrating the
% LLNCS macro package for Springer Computer Science proceedings;
% Version 2.21 of 2022/01/12
%
\documentclass[runningheads]{llncs}
%
\usepackage[T1]{fontenc}
% T1 fonts will be used to generate the final print and online PDFs,
% so please use T1 fonts in your manuscript whenever possible.
% Other font encondings may result in incorrect characters.
%
\usepackage{graphicx}
\usepackage{listings}
\usepackage{minted}
\usepackage{hyperref}

@@

Finally tagless style provides extensible, user friendly DSLs.

@@

One solution to this issue is to integrate container types such as tuples and
lists into the DSL language. However doing this eliminates the advantage of
having an embedded language. Manipulating tuple values will be cumbersome,
constantly requiring calls to custom implementations of
\mintinline{haskell}{fst}, \mintinline{haskell}{snd} etc. And for lists you'll
lose access to built-in Haskell list functionality.

@@

We need to make sure we don't attempt to insert the same substitution for two
different ASTs. Unfortunately, if there is a collision there's no way to escape
the state monad to prevent or modify the substitution. In the toy example,
compilation crashes, but we could catch an exception instead. Either way it's up
to the DSL writer to ensure they don't reuse the same label as a substitution.

@@

We profiled \mintinline{haskell}{cos} using both Kiselyov's method (i.e., hash-consing) and are own
(i.e., trie-consing with some explicit sharing). It wasn't possible to use
Kiselyov's explicit sharing method where meaningful so our code was able to
achieve greater performance, with only a small increase in memory consumption
(see Fig.~\ref{fig:hashvscons}).

\begin{figure}[!h]
  \centering
  \includegraphics[width=0.6\textwidth]{figs/cos_profile_hashcons_cropped.png}
  \includegraphics[width=0.6\textwidth]{figs/cos_profile_triecons_cropped.png}
  \caption{Comparison of cos using hash-cons vs trie-cons}
  \label{fig:hashvscons}
\end{figure}

@@

\begin{figure}[!h]
  \centering
  \includegraphics[width=0.6\textwidth]{figs/noexplicit_cropped.png}
  \hfill
  \includegraphics[width=0.6\textwidth]{figs/explicit_cropped.png}
  \caption{Comparison of tan profiling using no explicit sharing vs explicit}
  \label{fig:explicitvsno}
\end{figure}

>>>>

% This is samplepaper.tex, a sample chapter demonstrating the
% LLNCS macro package for Springer Computer Science proceedings;
% Version 2.21 of 2022/01/12
%
\documentclass[runningheads]{llncs}
%
\usepackage[T1]{fontenc}
% T1 fonts will be used to generate the final print and online PDFs,
% so please use T1 fonts in your manuscript whenever possible.
% Other font encondings may result in incorrect characters.
%
\usepackage{graphicx}
\usepackage{listings}
\usepackage{minted}
\usepackage{hyperref}
\usepackage{float}

@@

Finally tagless style provides extensible, user friendly DSLs. By providing one
interface for a variety of functionality, you can choose between shallow and
deep embeddings (see \cite{jovanovic2014yin} and
\cite{svenningsson2013combining}) and can easily extend the DSL without having
to alter the original class definition.

@@

One solution to this issue is to integrate container types such as tuples and
lists into the DSL language. However doing this eliminates the advantage of
having an embedded language. Manipulating tuple values will be cumbersome,
constantly requiring calls to custom implementations of
\mintinline{haskell}{fst}, \mintinline{haskell}{snd} etc. And for lists you'll
lose access to built-in Haskell list functionality. To get the full advantage of
embedding a DSL it should not only inherit the host language's syntactic and
semantic structure, but also it's programming environment, as is argued in
\cite{clements2001little}.

@@

We need to make sure we don't attempt to insert the same substitution for two
different ASTs. Unfortunately, if there is a collision there's no way to escape
the state monad to prevent or modify the substitution. In the toy example,
compilation crashes, but we could catch an exception instead if we used the
\mintinline{haskell}{ExceptT} transformer instead of a simple state monad.
Either way it's up to the DSL user to ensure they don't reuse the same label
as a substitution.

@@

We profiled \mintinline{haskell}{cos} using both Kiselyov's method (i.e.,
hash-consing) and are own (i.e., trie-consing with some explicit sharing). It
wasn't possible to use Kiselyov's explicit sharing method where meaningful so
our code was able to achieve greater performance, 154\% speedup, with only a
small increase in memory consumption (see Fig.~\ref{fig:hashvscons}).

\begin{figure}[H]
  \centering
  \includegraphics[width=0.6\textwidth]{figs/cos_profile_hashcons_cropped.png}
  \includegraphics[width=0.6\textwidth]{figs/cos_profile_triecons_cropped.png}
  \caption{Comparison of cos using hash-cons vs trie-cons}
  \label{fig:hashvscons}
\end{figure}

@@

\begin{figure}[H]
  \centering
  \includegraphics[width=0.6\textwidth]{figs/noexplicit_cropped.png}
  \hfill
  \includegraphics[width=0.6\textwidth]{figs/explicit_cropped.png}
  \caption{Comparison of tan profiling using no explicit sharing vs explicit}
  \label{fig:explicitvsno}
\end{figure}
